Probabilistic Topic Modelling for Spatial Smoothing in Mass Spectrometry Imaging

### 2016-12-11 docs/MSci_proposal/mprop.tex
The start of the initial design write-up.

% Intro
% - Literature review shows the mathematical underlying
% - Extract the relevant bits
\par Throughout the literature review, we have looked into the papers which are considered to have a significant influence on the LDA-like modelling. For this reason, we have set up the basis of the LDA-like model mathematical underlying. However, the approach in utilising those models in MSI like data sets have not been reviewed. Nevertheless, two sources of relevant research have been mentioned (the paper by Hooft et al. \cite{hooft} and the survey by Alonso et al. \cite{alonso-et-al}). Ultimately, this section is focused to emphasise the superior approaches of the previously reviewed papers.

% Generative processess
% - Comparison between static and dynamic data sets
\par The reviewed generative processes sets an assumption on the formulation of the data. We have reviewed two generative processes: static and dynamic. The static generative process is given by the initial LDA paper -- `Latent Dirichlet Allocation', and the dynamic generative processes is given by the 'Dynamic Topic Models' paper. These generative processes are expressed respectively in Figure \ref{fig:static} and Figure \ref{fig:dynamic} below. 
\begin{figure}[H]
  \centering
  \includegraphics[width=0.7\textwidth]{static}
  \caption{Static generative process.}
  \label{fig:static}
\end{figure}
\begin{figure}[H]
  \centering
  \includegraphics[width=0.7\textwidth]{dynamic}
  \caption{Dynamic generative process.}
  \label{fig:dynamic}
\end{figure}
In the previous plots, we have visualised the occurrences of a particular entity within a corpus. The static generative process suggests a random distribution of the entities, whereas the dynamic generative process induces a particular pattern. Since in metabolomics we expect the metabolites of same type to form clusters (rather than to be randomly distributed across the sample), the static generative process is more appropriate to reflect the nature of a biological entity development.

@@

The initial dynamic topic modelling motivation suggests the application of time-series into the generative process of data sets. However, the way we capture the ions is done simultaneously for the whole corpus. Therefore, we are relaxing the assumption of time-series. Instead of that, we will use the dynamic topic modelling techniques to establish the relationship between the time-slices (even though we are not taking time into account, we will stick to the same terminology to preserve the consistency). For example, if we assume that the data is generated in a 1-dimensional manner, then time-slice $t$ would depends on its adjacent time-slices $t-1$ and $t+1$. However, since the data sets we are working on are 2-dimensional, the induced relationship considers more adjacent time-slices. This relationship is shown in Figure \ref{fig:2-d_generative_process} below.
\begin{figure}[H]
  \centering
  \includegraphics[width=0.5\textwidth]{2-d_generative_process}
  \caption{2-dimensional generative process.}
  \label{fig:2-d_generative_process}
\end{figure}
In the boundary case shown on the left, time-slice $t_{0,0}$ relates to  three surrounding time-slices, whereas, in the general case indicated on the right side, time-slice $t_{1,1}$ relates to eight surrounding time-slices.

time-slice $t$ would be dependant on it's adjacent time-slices as shown  
Even though the dynamic topic modelling suggests

@@

% Design
% - The required model supplements
% -- Dictionary
\par
% To-do: induce time-series in terms of z axis

% Implementation
\par

>>>>

% Intro
% - Literature review shows the mathematical underlying
% - Extract the relevant bits
\par Throughout the literature review, we have looked into the papers establishing the mathematical underlying of the LDA-like models. However, the approach in utilising those models in MSI like data sets have not been reviewed. Nevertheless, two sources of relevant research have been provided (the paper by Hooft et al. \cite{hooft} and the survey by Alonso et al. \cite{alonso-et-al}). Ultimately, this section is focused to emphasise the superior approaches of the previously reviewed papers and establish the appropriate techniques in the generative and inference processes.

% Generative processess
% - Comparison between static and dynamic data sets
\par The reviewed generative processes sets an assumption on the formulation of the data. We have reviewed two generative processes: static and dynamic. The static generative process is given by the initial LDA paper -- `Latent Dirichlet Allocation', and the dynamic generative processes is given by the `Dynamic Topic Models' paper. These generative processes are expressed respectively in Figure \ref{fig:static} and Figure \ref{fig:dynamic} below. 
\begin{figure}[H]
  \centering
  \includegraphics[width=0.7\textwidth]{static}
  \caption{Static generative process.}
  \label{fig:static}
\end{figure}
\begin{figure}[H]
  \centering
  \includegraphics[width=0.7\textwidth]{dynamic}
  \caption{Dynamic generative process.}
  \label{fig:dynamic}
\end{figure}
In the previous plots, we have visualised the occurrences of a particular entity within a corpus. The static generative process suggests a random distribution of the entities, whereas the dynamic generative process induces a particular pattern. Since in metabolomics we expect the metabolites of same type to form clusters (rather than to be randomly distributed across the sample), the static generative process is more appropriate to reflect the nature of a biological entity development.

@@

The initial dynamic topic modelling motivation suggests the application of time-series into the generative process of data sets. However, the way we capture the ions is done simultaneously for the whole corpus. Therefore, we are relaxing the assumption of time-series. Instead of that, we will use the dynamic topic modelling techniques to establish the relationship between the time-slices (even though we are not taking time into account, we will stick to the same terminology to preserve the consistency). For example, if we assume that the data is generated in a 1-dimensional manner, then time-slice $t$ would depends on its adjacent time-slices $t-1$ and $t+1$. However, since the data sets we are working on are 2-dimensional, the induced relationship considers more adjacent time-slices. This relationship is shown in Figure \ref{fig:2-d_generative_process} below.
\begin{figure}[H]
  \centering
  \includegraphics[width=0.5\textwidth]{2-d_generative_process}
  \caption{2-dimensional generative process.}
  \label{fig:2-d_generative_process}
\end{figure}
In the boundary case shown on the left, time-slice $t_{0,0}$ relates to  three surrounding time-slices, whereas, in the general case indicated on the right side, time-slice $t_{1,1}$ relates to eight surrounding time-slices. Thereby, the notion of dynamic topic modelling is preserved even upon relaxing the assumption of time-series.

% Inference
% - Stochastic and optimisation based methods
% - Evaluation
\par The reviewed inference techniques have covered the mathematical underlying of two approaches: stochastic (sampling-based) and optimisation (variational). The variational methods are discussed in the `Latent Dirichlet Allocation' and `Dynamic Topic Models' papers, and the sampling-based method is discussed in the `Finding Scientific Topics' paper. Since both methods have been utilised in the state-of-the-art topic modelling applications, we will implement both of them in order to set up a discussion for the developed model's evaluation. However, since the authors of `Dynamic Topic Models' has utilised only the variational methods, we have to address the compatibility of the sampling-based methods in terms of applicability to the dynamic topic modelling.

@@

% Intro
% - The required model supplements
\par By covering the design of the model, we will set up the requirements for the project's execution and propose methods which would supplement the techniques derived from the reviewed literature. To start with, we will address the dimensionality reduction of the MSI data sets. Then, we will introduce the additional data capturing requirements shifting the set-up of the generative process. Finally, we will suggest an approach for utilising sampling-based methodology in the dynamic topic modelling.

% Dimensionality Reduction
% - Dictionary
% - Preprocessing
% -- Dismissing low intensity values
% -- Storing the ranges of ions
\par In order to address the scalability of the model in terms of the large MSI data sets, we will address two techniques for the dimensionality reduction: utilisation of a dictionary and pre-processing of the data. Both of these techniques reduces the sparsity of the data sets. To start with, the use of dictionary will eliminate the irrelevant use of memory in storing the words which do not occur in particular documents. For example, let's assume that the masses of the ions in our data can be represented as the set $\{1, 2, 3, 4, 5\}$, also assume that the ions can be represented in the form of (mass, intensity) tupple. Then, if our sample contained two following documents: $\mathbf{w_0} = \{(1, 100), (2, 200)\}$ and $\mathbf{w_1} = \{(3, 300)\}$, the sample representation in matrix and dictionary forms would be interchangeable with the expression given in Table \ref{tab:m_and_d} below.

\begin{table}[H]
\begin{center}
\begin{tabular}{|c||c|c|c|c|c|}
\hline
&1&2&3&4&5\\
\hline
$\mathbf{w_0}$&100&200&0&0&0\\
$\mathbf{w_1}$&0&0&300&0&0\\
\hline
\end{tabular}
\quad
\begin{tabular}{|c||c|c|c|c|c|}
\hline
&1&2&3&4&5\\
\hline
$\mathbf{w_0}$&100&200&$\O$&$\O$&$\O$\\
$\mathbf{w_1}$&$\O$&$$\O$$&300&$\O$&$\O$\\
\hline
\end{tabular}
\end{center}
\caption{Matrix and dictionary representation of the data set.}
\label{tab:m_and_d}
\end{table}

% Generative process
% - induce time-series in terms of z axis
\par

% Inference
\par
